---
title: "Floods and Tornados: Top Culprits for Human and Economic Perils"
author: "Chi Lam"
date: "September 23, 2015"
output: pdf_document
---
## Synopsis
Analyzing the NOAA Storm Database, we conclude that in the Unied States, torandos, excessive heat, and flash floods cause the most deaths; torandos, thunderstorm winds, and floods cause the most injuries; and floods, hurricane typhoons, and torandos cause the most economic losses.  Note that floods and tornados are the top causes for all 3 catergories.  


## Data Processing
This study uses the R software package to analyze the NOAA Storm Database.  The code blocks below documents all the R code generating the statistics and results.

### Libraries needed
```{r libs, message = FALSE, results=FALSE}
library(dplyr)
library(ggplot2)
```

### Loading data
```{r loadNOrg, cache=TRUE, message = FALSE}

#loading data
stormStat <- read.csv("repdata-data-StormData.csv.bz2")
```
Note that this process takes many minutes to complete, since the NOAA Storm Database is quite large.

### Statistics Generation
The following code block parses the costs of property damage, and crop damage, and combines them into the economic damage.  The sum of fatalities, injuries, and economic damage, and the frequency are also generated for each event.
```{r dataProc}
#interpret the prop and crop damage exponents
ltrexp <- c("", " ", "-", "?", "+", "0", "1", "2", "3", "4", 
	"5", "6", "7", "8", "B", "h", "H", "k", "K", "m", "M")
num10pwr <- c(0, 0, 0, 0, 0, 0, 1, 2, 3, 4,
	5, 6, 7, 8, 9, 2, 2, 3, 3, 6, 6)
stormStat$PROPDMGTTL <- stormStat$PROPDMG * 10^( num10pwr[match(stormStat$PROPDMGEXP, ltrexp)] )
stormStat$CROPDMGTTL <- stormStat$CROPDMG * 10^( num10pwr[match(stormStat$CROPDMGEXP, ltrexp)] )

#combining deaths with injuries, and property with crop damages
stormStat$humanHarm <- stormStat$FATALITIES * 50 + stormStat$INJURIES
stormStat$econDmg <- stormStat$PROPDMGTTL + stormStat$CROPDMGTTL

#generate the stats
stormStatSum <- stormStat %>% group_by(EVTYPE) %>% 
    summarize(frequency=n(), fatalities=sum(FATALITIES), injuries=sum(INJURIES), 
              humanHarm=sum(humanHarm), propDmg=sum(PROPDMGTTL), cropDmg=sum(CROPDMGTTL),
              econDmg=sum(econDmg));
stormStatSumByFatal <- stormStatSum[order(-stormStatSum$fatalities), ]
```

## Results
### Fatalities
The code below generates a bar chart of the top 10 events that cause the most fatalities in the United States.  Note that torandos, excessive heat, and flash floods cause the most deaths in the United States.
```{r fatalStats}
stormStatSumByFatalIdx <- order(-stormStatSum$fatalities)
top10fatal <- head(stormStatSum[stormStatSumByFatalIdx, c("EVTYPE", "frequency", "fatalities")], n=10)
top10fatal$EVTYPE <- factor(top10fatal$EVTYPE, levels = top10fatal$EVTYPE)
qplot(y=fatalities,x=EVTYPE, fill=frequency, data=top10fatal, geom="bar", stat="identity") + coord_flip() + 
    labs(title = "Top 10 Events by Fatalities", y="Fatalities", x="Event Type")
```

### Injuries
The code below generates a bar chart of the top 10 events that cause the most injuries in the United States.  Note that torandos, thunderstorm winds, and floods cause the most injuries in the United States.

```{r injuryStats}
stormStatSumByInjurIdx <- order(-stormStatSum$injuries)
top10injur <- head(stormStatSum[stormStatSumByInjurIdx, c("EVTYPE", "frequency", "injuries")], n=10)
top10injur$EVTYPE <- factor(top10injur$EVTYPE, levels = top10injur$EVTYPE)
qplot(y=injuries, x=EVTYPE, fill=frequency, data=top10injur, geom="bar", stat="identity") +     coord_flip() + 
    labs(title = "Top 10 Events by Injuries", y="Injuries", x="Event Type")
```

### Economic Losses
The code below generates a bar chart of the top 10 events that cause the most economic losses in the United States.  Note that floods, hurricane typhoons, and torandos cause the most economic losses in the United States.

```{r econStats}
stormStatSumByEconIdx <- order(-stormStatSum$econDmg)
top10loss <- head(stormStatSum[stormStatSumByEconIdx, c("EVTYPE", "frequency", "econDmg" )], n=10)
top10loss$EVTYPE <- factor(top10loss$EVTYPE, levels = top10loss$EVTYPE)
qplot(y=econDmg, x=EVTYPE, fill=frequency, data=top10loss, geom="bar", stat="identity") +     coord_flip() + 
    labs(title = "Top 10 Events by Economic Loss", y="Economic Loss(USD)", x="Event Type")
```

##Conclusion
 Analyzing the NOAA Storm Database, we conclude that in the Unied States, torandos, excessive heat, and flash floods cause the most deaths; torandos, thunderstorm winds, and floods cause the most injuries; and floods, hurricane typhoons, and torandos cause the most economic losses.  Note that floods and tornados are the top causes for all 3 catergories.  Policy makers should focus most of their efforts in mitigating the damages caused by floods and tornados.